Towards an Ecological Trait-data Standard

### 2018-05-01 manuscript/manuscript.Rmd
major revision of manuscript

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
- name: Malte Jochum \footnotemark[4]
- name: Ga\"etane Le Provost \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: Markus Fischer
- name: Pete Manning 
- name: Birgitta König-Ries
- name: Nadja K. Simons \footnotemark[6]
csl: amnat.csl
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07]. Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
For example, trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity and ecosystem functioning (refs). 
Using traits is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring both the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.
With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15].

<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. <!-- GLP: Maybe not the best example as they are not looking at trait diversty and species interactions ? 
Put instead:
Gravel et al. 2016 The meaning of functional trait composition of food webs for ecosystem functioning
And Coux et al. 2016 Linking species functional roles to their network roles
?
-->
<!--
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g. @reich97]. <!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].--> <!-- GLP: What do you mean by processes ? I’m not sure that the literature cited is always appropriate here ?
-->
<!--
Most importantly, the trait framework relates key ecosystem services for human well-being [@diaz11; @lavorel12; @allan15] back to changes in biodiversity in consequence to global change or local anthropogenic land use [ @diaz07; @gossner16; @mangels17; @jennings09; @simons16;  @mouillot13;  @vandewalle10;  @birkhofer15;@gamez-virues15; @birkhofer17]. 
--> 
<!-- CP: I like very much this whole paragraph (both writing and content), but I feel that it’s a bit out of the topic. It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important). I give some suggestions on how to cut it and rather develop more the following one, which people are surely less aware of and which sets more the aim of the paper. I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.

That’s the structure I try to follow:
1. Traits are important in ecology, evolution and conservation (i.e. this paragraph)

@@

4. One solution are databases. But these are costly and depend on the will of few persons/funding. They also cover few taxonomic groups. Matching between different databases is still difficult because of different definitions and structures. Most data are still uploaded separately.
We need a solution to make their use easy.

But of course it’s just a suggestion, feel free to happily ignore it!

I would choose 2 or 3 (also for the previous), anyway it’s impossible to be exhaustive. Maybe put (e.g. ref1, ref2) so make clear that it’s just examples

-->
<!-- GLP: 
The second paragraph is a bit confusing for me.

I wonder also if the introduction could be restructures like this :

Trait based approaches and their use
From the individual to population
Community level (assemblage, interactions, effects on ecosystem)
Macroecological question

 Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g. TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals). Say also something on the development of general protocols for measuring traits (e.g. Moretti et al. 2016; Freschet et al. 2017) ?

@@

There is just some copy-past to do. I didn’t change the structure of the introduction but if you like the structure that I proposed, we can discuss next week or I can help you with that. I like to write introduction 

@@

Huge number of references here- I think you can strip it down to a few key ones, and reviews. Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site. 
-->
<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, which continues to integrate a rapidly increasing number of fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->

However, fully exploiting this portential relies heavily on the broad availability and coverage of factual information and measurement data on all biotic taxa. <!-- PM:  Not 100% clear what you mean here- factual information could be anything- can you be more specific? -->
To mobilize data for this enterprise <!-- PM: Please define what the exercise is- too vague at present
-->, information must be made available digitally,  harmonized according to data standards and hosted in interlinked online resources. <!-- PM: I think this is a key justification- perhaps say what happens if we don’t do this – non repeatable work, differing definitions of traits that may lead to scientific flaws, repetition of the painstaking task of processing trait data e.g. to compute species level means. 
-->
The development of semantic-web standards [@wieczorek12; @page08] and ontologies [@walls12; @garnier17], as well as web-services [@kattge11; @parr16] for trait-data management is under way. <!-- CP: I felt that this needed a definition, not sure everybody know what an ontology is. Maybe we should guide more the reader that might not be familiar with the problems.
-->

Nonetheless, individual researchers often neglect the benefit of centralised databases <!-- PM: Explain what this benefit is. -->. Furthermore, many journals and research funding agencies require  data to be published without access restrictions [@swan12; @allison15]. However, this primary data is often posted on file hosting services, where no data standards are forced upon the uploaded material (e.g. Figshare, Dryad). The long-term outcome of this situation is an abundance of freely available data, but a lack of data standardisation.

In this paper, we (1.) discuss the importance of data standards for trait-based research and identify current deficits in standardisation from a pragmatic view of data providers and data users. We further (2.) review existing trait databases and web services, as well as initiatives for standardisation. 
We (3.) describe the common practice <!--PM: Current or your own new definition?--> for structuring trait data for ecological research. 
Based on these considerations, we (4.) propose a low-threshold <!-- PM: What’s this? Please define
-->  structure and vocabulary for describing trait-datasets compatible with existing ontologies <!-- PM & CAT: define--> . (5.) Finally, we provide an R-package that assists in aply this structure to own data and for the harmonisation of data from distributed sources.

With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards simple and low-threshold <!-- PM:Again, I’m not sure what this is a low threshold of? --> trait data standardisation.

Despite the obvious benefit and great potential of trait-based approaches, integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used. All of these factors render trait data extremely heterogeneous.

As a response to these issues, trait data have been standardized in centralized databases for specific organism groups and research questions.
<!-- PM: (refs, also give an e.g. or two for some big ones like TRY- saw there was a new one for tropical mammals a couple of days ago too and a huge Chinese plants one- in )--> 
These compiled data are either being published as static data tables or in the form of sophisticated online portals like the plant trait database TRY (see below for a review of current initiatives). These portals attract data submissions from a focal community of researchers and map heterogeneous data into a common scheme, and importantly also, provide networks for consensus building in measurement methods. Furthermore, database services offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets.

In contrast to standardized databases, open-access data publication in low-threshold repositories such as Figshare.com, Dryad (datadryad.org), Researchgate.net or Zenodo.org is becoming increasingly frequent. Such repositories foster decentralised data hosting with low thresholds regarding data standardisation and documentation.

Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets.
To cope with this trend towards distributed data sources, data brokering services (e.g. the German Federation on Biological Data; gfbio.org)  or database <!-- AO: Just “data” or “research data”--> management systems for scientific projects (e.g. BEXIS, XX <!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.

However, to mobilize the broad wealth of distributed data into structured databases <!-- AO: Maybe better: formats-->,  a certain minimum of information from original data providers is requrired, e.g. on the methodology and definitions of trait values or the covariates describing the ecological context of the trait observation. In order to get this accurate and well-defined input ecological researchers need to be aware of and understand basic principles of data standards and ontologies.  <!-- AO: The last paragraphs may need more sources/proofings?--><!-- PM: Literacy is quite a strong word- as illiterate is quite an insult. I think awareness is a problem too.  -->

 More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources.

@@

In this section, we review four kinds of initiatives that are of relevance for trait data standardisation: 
the first are projects that work with trait datasets that have been assembled out of a particular research interest, either by measurement or colated from literature; 
the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
the third aim at the standardization and develops consensus methods and definitions for traits by providing classifications and thesauri; 
the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct 'ontologies' for an application in a 'semantic web' of information.

We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the TOP Thesaurus of plant traits [@garnier17]. 
We show how the degree of trait-data standardization in exising datasets spans along this entire spectrum and which tools and standards do exist to acheive harmonization of data from multiple, distributed sources. 
The objective of this review is to raise awareness for the generic structure of trait data and aid ecological researchers to share and publish own datasets in the appropriate form.

@@

Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds in terms of metadata documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources on data management. 
In most of these datasets, the column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives. Datasets vary in the detail provided in the metadata, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a].

@@

Regarding open access, hardly any of these centralized databases for functional traits comply with the criteria demanded by journals and funding agencies for primary data publication. They incentivize data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.  
Time will show how these incentives fare with an increasing availability of open access trait data  on low-threshold data repositories. 
<!-- AO: In a sense of data availability to the public : yes. But do we discuss it here? We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs. non-open data, or?
bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->

@@

The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. To harmonize trait data across databases, future trait terminology initiatives should provide this functionality.
More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from different sources.

@@

Specifically designed for the context of plant traits, @kattge11 proposed a generic structure for plant trait data, that covers most potential use cases. This data structure is build around a central data table that contains observations, i.e. a single event of measurements on the same individual plant specimen at the same point in time. Identifiers link multiple measurements to the same observation, each measurement being well defined by additional standard tables. The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment. This structure can be implemented in any software or programming language for relational databases. 
<!-- FDS: enter comments of Jens Kattge here!-->

@@

However, these proposed standards and the available terms and extensions are responses to a demand from biodiversity data management for more structured input from the research community. 
None of the aforementioned aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standards. 
One reason for this is most certainly complexity: The data structures are designed for multi-layered databases rather than for static datasets for which a two-dimensional data table might suffice.  
If the traits are assembled for a specific research question with little ancillary information, it may be safer to add them as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road. This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet. 
<!-- DF: What is meant with “such information”? Does it still refer to the metadata in EML? In this case it should not be in a separate paragraph. -->
The trade-off is data-consistency and readability *vs.* avoidance of content duplication.

<!--When traits are assessed on a higher level of organisation, e.g. microbial functional traits assessed at the community level taken from a soil sample, key data columns would be placed in the same table as the core data are traits. Here, location or measurement information are the primary focus of the investigation and must be contained in the . 
<!-- PM: I get lost here-->
<!--A general recommendation, whether a column describes ancillary data or is part of the central dataset is ill advised. Therefore, our glossary of terms and its extensions should be used to describe the scientific data according to the study context. -->
<!-- FDS: shorten this and refer to existing standards for database use. the star-structure of ABCD, Kattge et al , and Garnier .-->

<!--while the existing trait data standards may cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots; PM: Vague- not sure what is missing here- be more specific --> 
The other reason is lack of awareness and literacy: many trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off, which is not affordable for small and intermediate size research projects.
By filling this gap, data brokering services (e.g. the German Federation on for Biological Data; gfbio.org) or database management systems for large-scale scientific projects (e.g. BExIS, NEON ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into consistent and interlinked data infrastructures. One goal of such initiatives is to facilitate the publication and standardization of data for researchers, e.g. by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses.<!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->

The practical importance of these frameworks for data providers and data users is often not obvious and not appropriately considered when storing or uploading data to low-threshold  file-publishing services. 
<!-- PM: There is also little incentive for researchers to do this- why bother? Incentives are also needed. -->
The pay-off is only playing out in the long run. Citations of trait datasets are higher when they are easier to access and relate to other data due to standard structures and vocabularies. 
Along with the push towards open access publication of primary data, the scientific community must come up with ways to incentivize and require data standardization.

@@

<!-- AO: You argue somehow that matrix format I a result of lit. review and so on. I would not agree with it. Your example also “just” points to aggregated data which menas one row is on species (aggregated) value.
Your example is (for me) not really a matrix. In a matrix you have a relation between two variables. In the first column and and on the upper row normally the values of one variable are written. An example would be species x body length. Then for body_length some classifications are written like (0-1, 1-2, 2-3,…) Then in the matrix the observation/abundance/… is noted. E.g. spec1 x “0-1” = 5, spec1 x “1-2” = 3, and so on. Your wexample hat already different variables (=traits) with different units as columns. That is a meaningfull representation of variables and values in a normal data table. Maybe it is good her to show the enormous number of different investigated traits and point to the resulting big (many columns = wide) tables.
--><!--
If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait wide-table format (Fig. 2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available. 
This format needs to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g. '999'). It may store additional information (e.g. on variation of means or literature source) in secondary columns or accompanying datasheets. 
The wide-table format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon). <!-- AO: You say it already 2 sentences before --><!--
Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14; @taugourdeau14] and using Bayesian approaches [@schrodt15]. <!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16]. --><!--AO: Why not? I can filter for a specific species and get all respective rows. But this is not a result a using a data format, it is a result of the data itself, FDS: no, since only one row per species exists; true that data itself don't cover this variation. -->
<!--

@@

>>>>

---
title: "Towards an Ecological Trait-data Standard"
author:
- name: Florian D. Schneider \footnotemark[1] \footnotemark[2]
- name: Malte Jochum \footnotemark[4]
- name: Gaetane Le Provost \footnotemark[2]
- name: Andreas Ostrowski \footnotemark[5]
- name: Caterina Penone \footnotemark[4]
- name: Markus Fischer \footnotemark[4]\footnotemark[2]
- name: Pete Manning \footnotemark[2]
- name: Birgitta König-Ries\footnotemark[5]
- name: Nadja K. Simons \footnotemark[6]
csl: amnat.csl
fontsize: 10pt
geometry: margin=1in
output:
  pdf_document:
    fig_caption: yes
    fig_crop: no
    keep_tex: yes
    template: pretty.template
  html_document:
    fig_caption: yes
    fig_width: 4
    highlight: null
    keep_md: yes
    theme: united
  word_document:
    fig_caption: yes
    fig_width: 6
bibliography: div05.bib
---

@@

<!-- CP:  I think we should develop more on what are the current problems/limitations (quickly because it’s then expanded in the paper) so it becomes evident that the standard+package fill a real need.

That’s the structure I try to follow:
1. Traits are important in ecology, evolution and conservation (i.e.  It is important to mention that traits are important in research and give some examples of questions that traits can help to solve but maybe not spend too long on it (most people know that they are important).

@@

4. One solution are databases. But these are costly and depend on the will of few persons/funding. They also cover few taxonomic groups. Matching between different databases is still difficult because of different definitions and structures. Most data are still uploaded separately.
We need a solution to make their use easy. 
-->
<!-- GLP: 
 Add a general paragraph on the fact that there have been some initiatives for the development of trait data base for plants (e.g. TRY), why is it important (provide quick access to plant trait values, allowing comparisons even between ecosystems and biomes), now the development of trait data base for other organisms (from microbes to mammals). Say also something on the development of general protocols for measuring traits (e.g. Moretti et al. 2016; Freschet et al. 2017) ?

@@

@@

Huge number of references here- I think you can strip it down to a few key ones, and reviews. Overall I think this section is quite a lot more complex than it needs to be – I think all you need to say is that it’s very popular, general and mechanistic and so adopted throughout eco and evo, with no sign of slowing down- the TRY database download stats can provide a quantitative measure of this- see their site. 
-->

<!-- 1. Traits are important in ecology, evolution and conservation -->

Functional traits are phenotypic characteristics that are related to the fitness and performance of an organism [@mcgill06d; @violle07]. Because trait-based approaches to ecological and evolutionary research questions allow for both mechanistic insight (refs) and generality (refs) recent years have seen a proliferation of trait-based research in a wide range of fields.
Trait-based studies have spanned the evolutionary basis of individual-level properties (refs) to global patterns of biodiversity [@diaz16] and ecosystem functioning (refs). 
For example, using traits is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms (refs), since they allow inferring the ecological role of organisms [@duarte11; @schrodt15] from their apparent features.

With this mechanistic depth, the trait framework relates losses of ecosystem function to changes in biodiversity, and on to climate or local anthropogenic land use [@wood15; @bello10; @diaz07; @lavorel12; @allan15; @diaz11; @jennings09; @hopfenmuller14; @mouillot13;  @vandewalle10; @perovic15].

<!-- Comparative studies of traits inform researchers about physiological and evolutionary trade-offs, and help identifying economic strategies of organisms beyond taxonomy [@menezes10; @forister15;  @siefert15; @gross09; @wright04]. Inferring an unknown functional role or ecological function of individuals from their apparent features [@duarte11; @schrodt15] is a promising venue to bypass taxonomic impediment, i.e. the fact that a majority of species are yet undescribed and little is known of their interactions with the environment and other organisms. 
Mutual 'trait-matching' of co-evolved organisms enables the inference of ecological interactions from observed features of live specimens [@leprovost17; @kalinkat13; @dehling14a] which may be used to inform ecological networks [@weiner14; @miranda13; @digel11; @schleuning15] and parametrise ecological population-dynamic models [@schneider12; @kalinkat13]. These analyses are also a promising venue for a multitrophic community ecology aiming to unravel dependencies and synchrony in trait spaces between interacting functional groups [@soliveres16]. <!-- GLP: Maybe not the best example as they are not looking at trait diversty and species interactions ? 
Put instead:
Gravel et al. 2016 The meaning of functional trait composition of food webs for ecosystem functioning
And Coux et al. 2016 Linking species functional roles to their network roles
?
-->
<!--
Since traits are so closely related to ecological mechanisms, they have proven useful for scaling individual level physiology and behaviour to functional diversity  of communities [@laliberte10a; @petchey09; @villeger08a; @hillerislambers12;@diaz98; @lavorel02], processes at the ecosystem scale [@fontaine05; @simons16; @gossner16; @moretti13;  @kuhsel15;  @birkhofer14], as well as macro-patterns of biodiversity at the global scale [@diaz16; e.g. @reich97]. <!-- Traits can be used to predict extinction risk [@fritz10; @pearson14] and susceptibility to environmental toxins [@rubach11].--> <!-- GLP: What do you mean by processes ? I’m not sure that the literature cited is always appropriate here ?
-->
<!--
Most importantly, the trait framework relates key ecosystem services for human well-being [@diaz11; @lavorel12; @allan15] back to changes in biodiversity in consequence to global change or local anthropogenic land use [ @diaz07; @gossner16; @mangels17; @jennings09; @simons16;  @mouillot13;  @vandewalle10;  @birkhofer15;@gamez-virues15; @birkhofer17]. 
--> 
<!-- In recent years, trait-based ecology has thus grown into a major field of ecological research, which continues to integrate a rapidly increasing number of fundamental and applied research questions. 
Given the expected future challenges in dealing with climate change and managing land use,  global trait synthesis is an interdisciplinary, cross-topic field with great potential. -->

<!-- 2. assemble trait data into databases -->

However, fully exploiting the potential of trait-based approaches relies heavily on the broad availability and taxonomic and regional coverage of trait data, both of present-day taxa ase well as in evolutionary deep-time. Researchers have a growing interest in assembling trait data from different sources since it pays off by getting answers to general questions. 
Integrating data from several sources across taxa, trophic levels, or organisational scales is difficult because these typically differ in numerous ways, including the identity of the taxa studied, the scale and place of study, methodology applied and taxonomic nomenclature used. All of these factors render trait data extremely heterogeneous.

<!-- 3. trait databases have been developed, with standard structure and terms --> 
To this end, trait data have been standardized in centralized databases for specific organism groups and regional scope, often centered around specific research questions (ref to TRY, panthera, etc.). 
These compiled data are either being published as static data tables that are the product of a one-time effort of collecting and harmonizing data (ref Amphibio, panthera, mammals), or in the form of sophisticated online portals, which attract data submissions from a focal community of researchers and provide research networks for consensus building in data resolution and measurement methods [e.g. @kattge11]. Furthermore, these portals map heterogeneous data into a common scheme, and importantly also, offer access control and data usage policies, to protect the rights of the original data providers, while simplifying data queries for synthesis researchers who seek to utilize these often-complex datasets. 
To enable the compatibility of data across data-bases, semantic-web standards [@wieczorek12; @page08] and ontologies of standard terms [@walls12; @garnier17] for trait-data are beeing developed.
Solving all these issues of data assemblage and harmonization in centralised infrastructures requires a huge effort in costs and personell (see below for a review of current initiatives).

<!-- 4. distributed trait data must be integrated --> 
Many journals and research funding agencies today require data to be published without access restrictions [@swan12; @allison15]. An increasing number of individual projects comply  with open data standards and publish their primary data on file hosting services like Figshare, Dryad or Researchgate, where no data standards are forced upon the uploaded material. The data standards mentioned above, that have been proposed for relational trait-databases, do hardly apply for these individual datasets.
Given this prospect, there will be no lack of data, but a lack of data standardisation hampering efficient future use of published datasets.

In this paper, we (1.) review existing trait databases and online portals, as well as initiatives for standardisation. In this review, we discuss the current practice and the importance of data standards for trait-based research and identify current deficits in standardisation from a pragmatic view of data providers and data users. Based on these considerations, we (2.) propose a minimal structure and vocabulary for describing trait-datasets, compatible with existing terminology standards for biodiversity data. Finally, we (3.)  propose the development of software that assists in aplying a harmonisation of data from distributed sources.

With this easy-to-use terminology and toolset, we hope to inform trait-data providers and trait-data users about the general importance of data standardisation and sketch a roadmap towards an accessible trait data standardisation.

<!-- FDS: Why 'ecological' trait data? what is special about this subject? Environmental context, community context, space and time, attributes of functional role in the ecosystem and energetic demands, interactions (not covered right now?), correlation and economic trade-offs in trait evolution -->

@@

In this section, we review four kinds of initiatives that are of relevance for trait data standardisation: 
- The first are projects that work with trait datasets which have been assembled out of a particular research interest, either by measurement or collated from literature; 
- the second aim to harmonize trait data from the literature or from direct measurement into trait databases and make those data widely available; 
- the third aim at the standardization and develops consensus methods and definitions for traits and provides standard terminologies; 
- the fourth kind of initiative feed the needs of biodiversity informatics and provide database structures and relational definitions to construct relational 'ontologies' for an application in a 'semantic web' of information.

We discuss these initiatives seperately although often they are developed in conjunction to serve a particular database project, as for instance in the case of the TRY plant database [@kattge11; @kattge11a] and the  Thesaurus of Plant Traits [TOP; @garnier17]. 
We show how the degree of trait-data standardization in exising datasets spans along this entire spectrum and which tools and standards do exist to acheive harmonization of data from multiple, distributed sources. 
The objective of this review is to raise awareness for the generic structure of trait data and aid ecological researchers to share and publish own datasets in the appropriate form.

@@

Most importantly, these platforms offer low-cost or free hosting of data with very low thresholds in terms of metadata documentation and standardisation, which make them attractive for small and intermediate sized research projects that cannot dedicate extra resources on data management. 
As a result of this, datasets also greatly vary in structure. 
The column descriptions and applied terminologies for taxa and traits are project specific, and rarely chosen to relate the data to other existing work or feed into larger database initiatives. Datasets vary in the detail provided in the metadata, e.g. for documenting descriptions of variables, measurement procedures or sampling context [@kattge11a]. 
If factual trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated quantitative measurements averaged from raw data, they usually are reported in a species$\times$trait wide-table format. Here, each row contains a species (or taxon) for which multiple traits are reported in columns. 
Similarly, when reporting raw data, researchers place observations of individual specimens in rows with multiple trait measurements applied to the same specimen placed in columns. 
The variability of the number and meaning of columns in these tables requires manual adjustments when merging multiple datasets.

<!-- Fig 1 here? -->

@@

Regarding open access, hardly any of these centralised databases for traits comply with the criteria demanded by journals and funding agencies for primary data publication. They incentivise data submissions by offering increased data visibility and usage, while providing data use policies that secure author attribution and, if demanded by the data owner, co-authorship.  
Time will show how these incentives fare with an increasing availability of open access trait data  on low-threshold data repositories. 
<!-- AO: In a sense of data availability to the public : yes. But do we discuss it here? We want to show how to make trait data sharing easier due to better data format, or?Not the topic of this article to discuss open vs. non-open data, or?
bKR: I am not quite sure what point you are trying to make it, the paragraph sort of goes back and forth between advantages and drawbacks of the different approaches. 
FDS: I think its part of the rationale for a simple data standard. More open data mean that we need more awareness for data standardisation. 
 -->

@@

To conclude, there is a shortage of globally valid trait thesauri or ontologies for traits that emerged from a standardisation effort of methodologies and broad consensus process within scientific field of ecology *and* that are sufficiently interlinked and accessible to fulfil semantic web standards. 
The benefit of such ontologies will increase if webservices and open Application Programming Interfaces (APIs) provide a way to extract the definitions and higher-level trait hierarchies programmatically via software tools. To harmonize trait data across databases, future trait terminology initiatives should provide this functionality.
Such thesauri would advance the possibilities of aggregated datasets and reduce noise and uncertainty. 
More compatibility and availability of trait thesauri could greatly improve the possibilities for a comparitive study of trait data from distributed sources.

@@

Specifically designed for the context of plant traits, @kattge11 proposed a generic structure for databases that covers most potential use cases of trait-based ecology. This data structure is build around a central data table that contains observations, i.e. a single event of measurements on the same individual plant specimen at the same point in time. This structure puts emphasis on the fact that trait  data are often measured for an analysis of correlations of multiple traits. Identifiers link multiple measurements (qualities) to the same observation (entity), each measurement being well defined by additional standard tables. The observations are also linked to a taxonomy and ancillary descriptors of the observation context, like location or experimental treatment. This structure can be implemented in any software or programming language for relational databases. 
<!-- FDS: enter comments of Jens Kattge here!-->

@@

These proposed standards and the available terms and extensions are responses to a demand from biodiversity data management for more structured input from the research community. 
Hovewer, hardly any of the aforementioned aggregated and harmonized datasets for birds, amphibians, or mammals employs such ontologies or semantic web standards.

One reason for this is most certainly complexity: The data structures are designed for multi-layered databases rather than for static datasets for which a two-dimensional data table might suffice.  
If the traits are assembled for a specific research question with few columns of ancillary information, it may be safer to add this information as extra columns to the two-dimensional data table, thus facilitating an analysis of correlations between traits and co-factors further down the road. This way, if datasets of different source are merged, the related information is readily available without the risk of breaking the reference to an external datasheet.

The other reason is lack of awareness of data providers for trait data standardization: many trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without clear incentive or immediate pay-off. Investing time and energy into research for the best and most sustainable data structure is not affordable for small and intermediate-size research projects.

By filling this gap, data brokering services (e.g. the German Federation on for Biological Data; gfbio.org) or data management systems for scientific projects (e.g. BExIS, NEON<!--BKR: Diversity Workbench (developed at SNSB), Metacat (https://knb.ecoinformatics.org/)--> ) are likely to gain importance. These services simplify and direct the standardized upload of research data and descriptive metadata into reliable and interlinked data infrastructures. One goal of such initiatives is to facilitate data publications and standardization for researchers, for instance by providing terminologies and ontologies for biodiversity data, and by consulting on publication licenses. <!-- AG: Other examples: Diversity Workbench (diversityworkbench.net), EDIT Platform for Cybertaxonomy (cybertaxonomy) -->

@@

<!-- AO: You argue somehow that matrix format I a result of lit. review and so on. I would not agree with it. Your example also “just” points to aggregated data which menas one row is on species (aggregated) value.
Your example is (for me) not really a matrix. In a matrix you have a relation between two variables. In the first column and and on the upper row normally the values of one variable are written. An example would be species x body length. Then for body_length some classifications are written like (0-1, 1-2, 2-3,…) Then in the matrix the observation/abundance/… is noted. E.g. spec1 x “0-1” = 5, spec1 x “1-2” = 3, and so on. Your wexample hat already different variables (=traits) with different units as columns. That is a meaningfull representation of variables and values in a normal data table. Maybe it is good her to show the enormous number of different investigated traits and point to the resulting big (many columns = wide) tables.
--><!--
If trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait wide-table format (Fig. 2*a*), with a column of values for each trait recorded and a row for each species (or taxon) for which data were available. 
The wide-table format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges of trait values for this taxon). <!-- AO: You say it already 2 sentences before --><!--
This format needs to clearly differentiate missing data and traits not applicable to a given taxon (often filled in as 'NA' or dummy numbers, e.g. '999'). It may store additional information (e.g. on variation of means or literature source) in secondary columns or accompanying datasheets.

Quantitative traits reported at this resolution are often used to calculate community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values of a rarely described species, may be inferred from these tables when combined with phylogenies ['gap filling', @swenson14; @penone14; @taugourdeau14] and using Bayesian approaches [@schrodt15]. <!--However, these matrices do not allow estimates of within-species variation or assessments of the quality of the reported measurement or fact [@guralnick16]. --><!--AO: Why not? I can filter for a specific species and get all respective rows. But this is not a result a using a data format, it is a result of the data itself, FDS: no, since only one row per species exists; true that data itself don't cover this variation. -->
<!--

@@

<!--The trade-off is data-consistency and readability *vs.* avoidance of content duplication. -->
